---
title: "team"
author: "Patrick Wenzel, Kordell Schrock"
date: "10/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(tidyverse)
library(hash)
library(dplyr)
library(ggplot2)
library(devtools)
library(data.table)
library(anytime)
library(data.table)
library(knitr)
library(tidyr)
library(reshape2)
```

```{r}
library(readxl)
defense <- read_excel('./cyclonesFootball2019.xlsx', sheet='Defensive')
offense <- read_excel('./cyclonesFootball2019.xlsx', sheet='Offensive')
biography <- read_excel('./cyclonesFootball2019.xlsx', sheet='Biography')
str(defense)
```
## Part 1: Cleaning data
```{r}

offenseClean <- offense %>%
  mutate(Name = factor(Name),
         Opponent_Opponent = factor(Opponent_Opponent)) %>%
  mutate(across(Rushing_ATT:Passing_INT, as.numeric)) #Change to numeric
    #cleaning the offense

defenseClean <- defense %>%
  mutate(Name = factor(Name),
         Opponent_Opponent=factor(Opponent_Opponent)) %>%
  mutate(across(Tackles_Solo:Pass_PB, as.numeric)) #Change to numeric
  #cleaning the defense

biographyClean <- biography %>%
  mutate(Weight = as.numeric(Weight)) %>% # Change to numeric column
  separate(Height, c("feet","inches"), "-") %>% # create 2 columns feet and inches from Heights
  mutate(Height = as.numeric(feet)* 12 + as.numeric(inches)) %>% #get numeric inches value from the values of the 2 new columns feet*12 + inches
  select(-feet, -inches)
  #cleaning the biography

str(defenseClean) #print dataset
str(offenseClean) #print dataset
str(biographyClean) #print dataset

```
# Part 2: Tidying
## Reformat the defClean data frame into a tidy format using pivot_longer. The type of statistic (Tackles_Solo, Tackles_ASST, etc) is added as a new key column named stat.

```{r}
#1
defenseClean<-defenseClean %>%
  pivot_longer(Tackles_Solo:Pass_PB, names_to = 'stat')
```
## 2 Compare the distributions of the defensive statistics. What defensive skills are rare?
### 2: While looking at the BarChart we see that TurnOver_FF, TurnOver_FR, and TurnOver_INT are the rarest. Then looking into the Histogram we see again that TurnOver_FF, TurnOver_FR, and TurnOver_INT are the rarest. 

```{r}
#2
sumOfStats <- defenseClean %>% 
  group_by(stat) %>% #group by statistics
  summarise(sum = sum(value)) %>% #get the sum
  arrange(sum)

print(sumOfStats) #print the dataset

ggplot(defenseClean, aes(x=value)) + geom_histogram(bins=10) + facet_wrap(~stat) + ggtitle('Histograms of Defensive Stats bins=10') + xlab('Value of Defensive Stats') + ylab('Count of Defensive Stats')
#create histogram of Defensive Stats

ggplot(defenseClean, aes(x = reorder(stat, value), y = value)) + geom_bar(stat = 'identity') + coord_flip() + labs(x = 'defensive skills', y = 'Frequency', title = 'Defensive Stats Freq')
#create bar chart of Defensive Stats
```
## 3. Did ISU have better defense against Iowa or Notre Dame? Answer this question by creating a scatterplot with x- and y-axes being the number of solo tackles (of each player). A large number of solo tackles is an indicator of good defense.
 
### 3: Looking at the Scatter Plot Notre Dame seems to have higher counts of Solo Tackles per player. So Notre Dame must have had a better defence than Iowa. Also looking at the statistical data of aggregating the teams with solo tackles we see again Notre Dame beats Iowa in sum of Tackles and average tackles per player.
```{r}
#3
iowaVSnotreDame <- defenseClean %>%
  filter(stat == "Tackles_Solo" & (Opponent_Opponent == "Notre Dame" | Opponent_Opponent == "Iowa")) %>% #filter by Solo Tackles and team either Notre Dame or Iowa
  rename(teams = Opponent_Opponent) #rename column

str(iowaVSnotreDame) #display dataframe

tackles<- aggregate(iowaVSnotreDame$value ~ iowaVSnotreDame$teams , data=iowaVSnotreDame, sum) #aggregate tackles sum per team
averagetackles<- aggregate(iowaVSnotreDame$value ~ iowaVSnotreDame$teams , data=iowaVSnotreDame, mean)#aggregate tackles mean per team
print(tackles) #print Tackles
print(averagetackles) #print average Tackles
  
ggplot(iowaVSnotreDame, aes(x = value, y = Name)) + facet_wrap(iowaVSnotreDame$teams) + geom_point() + ggtitle('Solo Tackles for the Iowa and Notre Dame Game') + xlab('Number of Solo Tackles Per Player') + ylab(' Football Players') 
#Create scater plot of Iowa and Notre Dame Players for Solo Tackles for each player.
# Use facet_wrap() to to show teams variables: Iowa and Notre Dame
```
## 4. Separate the Hometown column into two columns, namely the city and the state. Print the first few rows of these two columns. (Hint: look at the sep= argument of separate)
```{r}
#4
biographyClean <- biographyClean %>%
  separate(Hometown, c('City', 'State'), sep =", ") #create 2 columns City and State from Hometown.
head(biographyClean) #head of clean biography dataset
```
## How many players are there from each state?
```{r}
#5
playersPerState <- biographyClean %>%
      group_by(State) %>% #group by state
      summarize(count = n()) %>% #tally
     
print(playersPerState) #print dataset
```
